A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-05-27 paper/writeup.tex
t[NULL] fixup

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, this returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$);
        these are NULL if $n$ has no parent.
        The variable $m_e$ will record the number of new edges to be added,
        and $m$ will record the number of extended haplotypes.
        Note that if no haplotypes can be extended ($m=0$), we return $m_e$ as $\infty$
        so that the path is not merged in \algorithmref{alg:extend}.
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{MergeNum}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
    	 %\tcc{We construct paths beginning at $c$ and traverse through inheritance in $T_O$ and $T_I$.}
        \text{Let } $p_i = P_I[c]$, 
        $t_i = t[p_i]$,
        $p_o = P_O[c]$,
        $t_o = t[p_o]$,
        \text{ and }$m_e = m = 0$.\;
        \While{\tn{True}}{
            \text{Set } $y_i = (p_i \neq NULL)
                \ \&\ (p_i \notin T_O)
                \ \&\ (t_i < t_o)$ \tcc*{Next node is from $T_I$.}
            \text{and } $y_o = (p_o \neq NULL)
                \ \&\ (p_o \notin T_I)
                \ \&\ (t_o < t_i)$ \tcc*{Next node is from $T_O$.}
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$}{
                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$c \to p_i$ is a new edge.}
                }
                \text{Set } $c = p_i$,
                $p_i = P_I[p_i]$, \text{ and}\;
                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
            }{
                \If{$P_O[c] \neq p_o$}{
                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$c \to p_o$ is a new edge.}
                }
                \text{Set }$c = p_o$,
                $p_o = P_O[p_o]$, \;
                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
                \text{and } $m = m + 1$ . \tcc*[r]{$c \to p_o$ would be extended.}
            }
        }
        \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            \text{Set } $m_e = \infty$. \tcc*[r]{The paths cannot be merged.}
        }
        \KwRet{$m_e$}\;
    }
\end{algorithm2e}

>>>>

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, this returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$);
        these are NULL if $n$ has no parent,
        and we take $t[\text{NULL}] = \infty$.
        The variable $m_e$ will record the number of new edges to be added,
        and $m$ will record the number of extended haplotypes.
        Note that if no haplotypes can be extended ($m=0$), we return $m_e$ as $\infty$
        so that the path is not merged in \algorithmref{alg:extend}.
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{MergeNum}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
    	 %\tcc{We construct paths beginning at $c$ and traverse through inheritance in $T_O$ and $T_I$.}
        \text{Let } $p_i = P_I[c]$, 
        $t_i = t[p_i]$,
        $p_o = P_O[c]$,
        $t_o = t[p_o]$,
        \text{ and }$m_e = m = 0$.\;
        \While{\tn{True}}{
            \text{Set } $y_i = (p_i \neq NULL)
                \ \&\ (p_i \notin T_O)
                \ \&\ (t_i < t_o)$ \tcc*{Next node is from $T_I$.}
            \text{and } $y_o = (p_o \neq NULL)
                \ \&\ (p_o \notin T_I)
                \ \&\ (t_o < t_i)$ \tcc*{Next node is from $T_O$.}
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$}{
                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$c \to p_i$ is a new edge.}
                }
                \text{Set } $c = p_i$,
                $p_i = P_I[p_i]$, \text{ and }
                $t_i = t[p_i]$ .\;
                % $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
            }{
                \If{$P_O[c] \neq p_o$}{
                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$c \to p_o$ is a new edge.}
                }
                \text{Set }$c = p_o$,
                $p_o = P_O[p_o]$,
                $t_o = t[p_o]$ \;
                % $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
                \text{and } $m = m + 1$ . \tcc*[r]{$c \to p_o$ would be extended.}
            }
        }
        \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            \text{Set } $m_e = \infty$. \tcc*[r]{The paths cannot be merged.}
        }
        \KwRet{$m_e$}\;
    }
\end{algorithm2e}
